Typed Functional Genetic Programming

### 2013-07-07 tex/thesis.tex
[tex]

@@

@@

Let $V$ be infinite countable set of {\it 
variable names}.  \\* 
Let $C$ be set of {\it constant names}.	 \\*		
Then $\Lambda$ is set of {\it \lterms} inductively defined as follows:

@@

@@

We use upper case symbols such as 
$M,N,M_1,M_2,...,N_1,N_2,...$
to denote arbitrary lambda terms in contrast to
those in lower case such as
$a,b,c,...,x,y,z,$
$f_1,f_2,...,g_1,g_2,...$
denoting variable names, elements of $V$.
Sometimes we will use lower case symbols
to denote arbitrary variable names (such as
$x$ in $\lamb{x}{M}$),
and at other times, we will use them
as specific variable names in specific terms 
(such as $x$ in $\lamb{x}{x}$).
The distinction between the two should be
clear from the context.\\

@@

Let us quickly define $ST(M)$ set of subterms of a term $M$:\\

\begin{easylist}[itemize]
& If $M = x$ then $ST(M) = {x}$.
& If $M = (P~Q)$ then $ST(M) = \{(P~Q)\} \cup ST(P) \cup ST(Q)$
& If $M = \lam{x}{M}$ then $ST(M) = \{\lam{x}{M}\} \cup ST(M)$
\end{easylist}~

\textbf{TODO FV a Substituci...}

@@

Any term of a form $\lamb{x}{M}N$ can be $\beta$-reduced to 
term $M[x:=N]$. This fact is written as relation of those
two terms in the following way:

$$\lamb{x}{M}N \rightarrow_\beta M[x:=N]$$

@@

\subsection{Reductions for $c \in C$}
TODO možná vynechat...

\subsection{Normal form}

\subsection{Long normal form}

\textbf{TODO} 
\begin{itemize} 
	\item TALK ABOUT "parenthesis" conventions 
	      (and packing of lambda abstractions).
	\item Popsat ve stručnosti beta redukci,
	      etha redukci, atd.
	      Ale hlavně promluvit krátce o intuici
	      a nějak to napojit na to jak to používám já
	      pak v tom GP.
	\item promluvit zde o long normal form což propojuje 
		  tuhle sekci se sekcí o barenregt-like 	
		  pravidlech \ref{barlike}.
\end{itemize}

@@

@@

As is said in \cite{barendregt10} there are two
inference/grammar rules which together generate
simply typed lambda terms in their long normal forms
(from section \ref{deflam}).

@@

Inference rule 2: 
\[
	\frac{ (f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma \qquad
	       \Gamma \vdash M_1 : \rho_1 \quad
	       \dotsm \quad
	       \Gamma \vdash M_m : \rho_m        
	      }
	     {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
Proof of correctness (\textbf{TODO REPAIR} Conceptually it is ok but there is sazba-bug somewhere): 
\[
   \dfrac
     {\dfrac
      {\dfrac
       {\dfrac         
         {\dfrac  
          {\dfrac
           {\boxed{(f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma}}
           {\Gamma \vdash f : \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha}
           \quad
           \boxed{\Gamma \vdash M_1 : \rho_1} }
          {\Gamma \vdash (f M_1) : \rho_2 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha }
          }{\vdots} 
         \quad 
         \ddots }
       {\Gamma \vdash (f M_1 \dots M_{m-2}) : \rho_{m-1} \rightarrow \rho_m \rightarrow \alpha}
       \quad
       \boxed{\Gamma \vdash M_{m-1} : \rho_{m-1}}  }
      {\Gamma \vdash (f M_1 \dots M_{m-1}) : \rho_m \rightarrow \alpha}       
      \quad 
      \boxed{\Gamma \vdash M_m : \rho_m} }
	 {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
... there is a grammar rule:
\[ 
	( \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]
\\\\
\textbf{TODO} 
\begin{itemize}
	\item SHOW correctness of those inference rules by composing them of 
		  $E^{\rightarrow}$, $I^{\rightarrow}$ and \textit{axiom}.
	\item SHOW more examples of inference rules transformed into grammar rules.
	\item DESCRIBE general algorithm for this transformation.
	\item TALK ABOUT $\tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$ 
	\item TALK ABOUT $\beta \eta$-normal form which is generated by this method.
\end{itemize}

@@

>>>>

%takle asi spíš psát ty anglický věty !!!!!!!!!!!!!!!!!!!!!! 
% takle nak to píše barendrecht :
%$\FV(M)$ the set of free variables of term M is defined inductively as follows.

@@

\usepackage{amssymb}

@@

@@

Let $V$ be infinite countable set of {\it 
variable names}.  \\* 
Let $C$ be set of {\it constant names}.	 \\*		
Then $\Lambda$ is set of {\it \lterms} defined inductively as follows
\footnote{Function application and $\lambda$-abstraction are concepts
well known from common programming languages. 
For example in JavaScript 
$(M~N)$ translates to expression \texttt{$M$($N$)} and
$\lamb{x}{M}$ translates to expression \texttt{function($x$)\{return $M$;\}}.
In other words, the $\lambda$-abstraction is equivalent to 
\textit{anonymous function}. 
}.

@@

We use upper case symbols such as 
$M,N,M_1,M_2,...,N_1,N_2,...$
to denote arbitrary lambda terms in contrast with
those in lower case such as
$a,b,c,...,x,y,z,$
$f_1,f_2,...,g_1,g_2,...$
denoting variable names, elements of $V$.
Sometimes we will use lower case symbols
to denote arbitrary variable names (such as
$x$ in $\lamb{x}{M}$),
and at other times, we will use them
as specific variable names in specific terms 
(such as $x$ in $\lamb{x}{x}$).
The distinction between the two should be
clear from the context.\\

@@

\newcommand{\ST}{\mathop{\mathrm{ST}}}
\newcommand{\FV}{\mathop{\mathrm{FV}}}

In order to demonstrate what is a subterm of a term
\lets define inductively $\ST(M)$ set of all subterms of a term $M$.
\begin{align*}
\ST(x)          &= \{x\} \\
\ST((P~Q))      &= \{(P~Q)\} \cup \ST(P) \cup \ST(Q) \\
\ST(\lam{x}{P}) &= \{\lam{x}{P}\} \cup \ST(P) 
\end{align*}

%takle asi spíš psát ty anglický věty !!!!!!!!!!!!!!!!!!!!!! 
% takle nak to píše barendrecht :

$\FV(M)$ the set of free variables of a term M is defined inductively as follows.
\begin{align*}
\FV(x)          &= \{x\} \\
\FV((P~Q))      &= \FV(P) \cup \FV(Q) \\
\FV(\lam{x}{P}) &= \FV(P) \smallsetminus \{x\}
\end{align*}

$M[x:=N]$ the substitution of a term $N$ for the free occurrences of 
a variable $x$ in a term $M$ is defined inductively as follows.
\begin{align*}
x[x:=N]           &= N \\
y[x:=N]           &= y \text{ ~~~~where } x \not= y  \\
(P~Q)[x:=N]       &= ( P[x:=N]  ~ Q[x:=N] ) \\
\lamb{x}{P}[x:=N] &= \lam{x}{P}\\
\lamb{y}{P}[x:=N] &= \lam{y}{( P[x:=N] )} \text{ ~~~~where } x \not= y
\end{align*}

\textbf{TODO možná zmínit že y nesmí být v N volně v připadě\\
$\lamb{y}{P}[x:=N] = \lam{y}{( P[x:=N] )}$ }, ale řekl bych že zbytečnej detail

@@

\newcommand{\bRedex}{$\beta$-redex\xspace}
\newcommand{\bRedexes}{$\beta$-redexes\xspace}
\newcommand{\bArrow}{\rightarrow_\beta\xspace}
\newcommand{\eArrow}{\rightarrow_\eta\xspace}
\newcommand{\eeArrow}{\rightarrow_{\eta^{-1}}\xspace}

A term of a form $\lamb{x}{M}N$ is called \textit{\bRedex}.
A \bRedex can be $\beta$-reduced to term $M[x:=N]$. 
This fact is written as \textit{relation} $\bArrow$ 
of those two terms:
\begin{equation} \label{eq:bRed}
\lamb{x}{M}N \bArrow M[x:=N]
\end{equation}
It is also possible to reduce \textit{subterm \bRedexes} 
which can be formally stated as:
\begin{align*}
P \bArrow Q &\then (R~P)      \bArrow (R~Q) \\
P \bArrow Q &\then (P~R)      \bArrow (Q~R) \\
P \bArrow Q &\then \lam{x}{P} \bArrow \lam{x}{Q}  
\end{align*}

@@

Similarly as for $\beta$-reduction we can define $\eta$-reduction 
except that instead of \ref{eq:bRed} we use:

$$\lamb{x}{(M~x)} \eArrow M \text{ ~~~~where } x \not\in FV(M) $$

\subsection{$\eta^{-1}$-reduction}

$\eta^{-1}$-reduction (also called $\eta$-expansion) is 
the reduction converse to $\eta$-reduction.
Again it may be obtained by replacing \ref{eq:bRed}, now with:

$$M \eeArrow \lamb{x}{(M~x)} \text{ ~~~~where } x \not\in FV(M) $$

\subsection{Reductions for $c \in C$}
\textbf{TODO možná vynechat...}

\subsection{$\beta$-normal form}

A \lterm is a \textit{$\beta$-normal form} if it does not have a $\beta$-redex as
subterm.
\\\\
A normal form may be thought of as a result of a term evaluation.

~\\\\
\textbf{TODO} 
\begin{itemize} 
	\item promluvit krátce o intuici
	      a nějak to napojit na to jak to používám já
	      pak v tom GP.
	\item Citovat nějakýho bárendrechta jako zdroj podrobnějších informací
\end{itemize}

@@

\subsection{Long normal form}
\label{lnf}

\textbf{TODO : promluvit zde o long normal form což propojuje 
		  tuhle sekci se sekcí o barenregt-like 	
		  pravidlech \ref{barlike}.
}

@@

As is said in \cite{barendregt10} there are two
inference/grammar rules which together generate
simply typed lambda terms in their long normal forms
(defined in \ref{lnf}).

@@

Inference rule 2: 
\[
	\frac{ (f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma \qquad
	       \Gamma \vdash M_1 : \rho_1 \quad
	       \dotsm \quad
	       \Gamma \vdash M_m : \rho_m        
	      }
	     {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
Proof of correctness (\textbf{TODO REPAIR} Conceptually it is ok but there is sazba-bug somewhere): 
\[
   \dfrac
     {\dfrac
      {\dfrac
       {\dfrac         
         {\dfrac  
          {\dfrac
           {\boxed{(f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma}}
           {\Gamma \vdash f : \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha}
           \quad
           \boxed{\Gamma \vdash M_1 : \rho_1} }
          {\Gamma \vdash (f M_1) : \rho_2 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha }
          }{\vdots} 
         \quad 
         \ddots }
       {\Gamma \vdash (f M_1 \dots M_{m-2}) : \rho_{m-1} \rightarrow \rho_m \rightarrow \alpha}
       \quad
       \boxed{\Gamma \vdash M_{m-1} : \rho_{m-1}}  }
      {\Gamma \vdash (f M_1 \dots M_{m-1}) : \rho_m \rightarrow \alpha}       
      \quad 
      \boxed{\Gamma \vdash M_m : \rho_m} }
	 {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
... there is a grammar rule:
\[ 
	( \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]
\\\\
\textbf{TODO} 
\begin{itemize}
	\item SHOW correctness of those inference rules by composing them of 
		  $E^{\rightarrow}$, $I^{\rightarrow}$ and \textit{axiom}.
	\item SHOW more examples of inference rules transformed into grammar rules.
	\item DESCRIBE general algorithm for this transformation.
	\item TALK ABOUT $\tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$ 
	\item TALK ABOUT $\beta \eta^{-1}$-normal form which is generated by this method.
\end{itemize}

@@

\textbf{TODO : Odiskutovat nějak tohle:
přirozeně chceme při produkci velkýho množství termů obhospodařovávat nějaký
velký množství rozdělanejch termů v nějaký struktuře a postuipně je dodělávat.
Proto přichází jako tahle struktura na mysl nějaká prioritní fronta,
otázka je kde vzít ty priority a A* je jednoduchá ale přitom optimum hledající 
odpověď na tuhle otázku, dostatečně volná skrz tu heuristiku.  }

\textbf{TODO: určitě někde zmínit ospravedlnění proč si napsat dokazovač sám
a né požít nějakej už hotovej - de o to, že dokazovače jsou delaný na to aby
udělali jeden důkaz typicky složitýho tvrzení, my sme v situaci kdy
chceme co nejvíc (a co možná co nejvíc rozmanitejch) důkazů jednoduchýho problému.
 }
